---
title: "Getting started with pagoda2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with pagoda2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
## The vignette needs the optional UMAP / Leiden / plotting backends. If any is
## missing (e.g. a stripped check environment) we show the code but do not run it.
have.deps <- all(vapply(c("uwot", "leidenAlg", "ggplot2"), requireNamespace,
                        logical(1), quietly = TRUE))
knitr::opts_chunk$set(eval = have.deps)
```

`pagoda2` analyzes single-cell count data: normalization, variance modeling,
dimensionality reduction, graph-based clustering, embeddings, and marker
detection. The pipeline is driven by an R6 `Pagoda2` object whose steps are
**generic verbs** — `runReduction()`, `runGraph()`, `runClustering()`,
`runEmbedding()`, `runMarkers()` — where the algorithm is selected by a
`method=` argument (there is no `runPCA`/`runLeiden`). `run()` chains the whole
standard sequence in one call.

Multi-modal data (CITE-seq, multiome) is supported through **facets**, one per
modality; ordinary single-modality analysis (below) never needs them. See the
tutorial notebooks linked from the
[package README](https://github.com/kharchenkolab/pagoda2) for CITE-seq and
ATAC/multiome workflows on real data.

This vignette uses a small **synthetic** dataset so it is fast and
self-contained; the commands are identical on real data.

## A toy dataset

We simulate 480 cells from four populations, each marked by its own block of 50
genes elevated over a shared background. The expected input is a sparse
gene-by-cell count matrix.

```{r}
library(pagoda2)
set.seed(42)

n.pop <- 4; cells.per <- 120; n.markers <- 50; n.background <- 600
population <- rep(seq_len(n.pop), each = cells.per)
n.cells <- length(population)
n.genes <- n.pop * n.markers + n.background

## per-gene, per-cell Poisson rate: low background, raised for a population's markers
rate <- matrix(0.5, n.genes, n.cells)
for (k in seq_len(n.pop)) {
  marker.rows <- ((k - 1) * n.markers + 1):(k * n.markers)
  rate[marker.rows, population == k] <- 4
}
rate <- sweep(rate, 2, runif(n.cells, 0.8, 1.2), "*")   # per-cell library-size variation

counts <- matrix(rpois(length(rate), rate), n.genes, n.cells)
rownames(counts) <- c(paste0("Marker", seq_len(n.pop * n.markers)),
                      paste0("Bg", seq_len(n.background)))
colnames(counts) <- paste0("cell", seq_len(n.cells))
counts <- as(Matrix::Matrix(counts, sparse = TRUE), "dgCMatrix")
dim(counts)   # genes x cells
```

## Build the object and run the standard workflow

```{r}
p2 <- Pagoda2$new(counts, log.scale = TRUE, min.cells.per.gene = 5,
                  min.transcripts.per.cell = 10, n.cores = 1, verbose = FALSE)

p2$run(plots = "none", verbose = FALSE,
       variance  = list(use.raw.variance = TRUE),
       pca       = list(nPcs = 15, n.odgenes = 300),
       graph     = list(k = 20),
       embedding = list(n_neighbors = 15, min_dist = 0.3),
       markers   = list(z.threshold = 2))
```

`run()` performed QC, variance modeling, PCA, a kNN graph, a UMAP embedding,
Leiden clustering, and marker detection, storing each result under a canonical
name. (Every step is also callable on its own, e.g.
`p2$runReduction(method = "pca")` or `p2$runClustering(method = "leiden")`.) The
recovered clustering separates the four populations:

```{r}
table(population = population, cluster = p2$cellMeta[[p2$defaultGrouping]])
```

## Embedding

`plotEmbedding()` draws the stored UMAP, colored by the default (Leiden)
grouping.

```{r, fig.width = 5, fig.height = 4.2}
p2$plotEmbedding(mark.groups = TRUE, size = 1, alpha = 0.6,
                 title = "Leiden clusters")
```

## Markers

The default run computed markers for the Leiden grouping. `getTopMarkers()`
returns the top genes per cluster as a data frame; each cluster recovers its
population's marker block.

```{r}
p2$getTopMarkers(n = 3)
```

The dot plot summarizes them — color is mean expression, dot size the fraction
of cells expressing the gene.

```{r, fig.width = 7, fig.height = 4.5}
p2$plotMarkerDotPlot(markers = p2$defaultGrouping, n.genes.per.group = 4,
                     order.groups = TRUE, dot.scale = 6)
```

## Next steps

Annotate the clusters from the marker evidence and make the annotation the
default grouping, after which every plot uses it:

```r
p2$annotateClusters(from = "leiden", to = "cell_type",
                    map = c("1" = "type A", "2" = "type B"), setDefault = TRUE)
```

For multi-modal data each modality is a **facet** and the same verbs take a
`facet=` (one modality) or `facets=` (joint) argument; see the CITE-seq and
ATAC/multiome notebooks linked from the
[README](https://github.com/kharchenkolab/pagoda2).
