---
title: "Pagoda2 processing with GO aspects"
author: "Nikolas Barkas"
date: "August 14, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries
```{r}
# Load libs
library(pagoda2)
library(Matrix)
library(nbHelpers)
````

Load 10X matrix
```{r cars}
path <- "path/to/tenXmatrices"
x <- read10xMatrix(path=path)
```

Perform filtering of matrices
```{r}
# filter cells based on the gene/molecule dependency
t.filter.for.valid.cells <- function(countMatrix,min.cell.size=500, max.cell.size=5e4,p.level=min(1e-3,1/ncol(countMatrix)),alpha=0.1,do.par=T) {
          if(do.par) { par(mfrow=c(1,2), mar = c(3.5,3.5,2.0,0.5), mgp = c(2,0.65,0), cex = 1.0);}
                  hist(log10(colSums(countMatrix)),col='wheat',xlab='log10[ molecules ]',main='')
                  # some of the cells are very large .. those can skew the analysis of more subtle populations (too much bias ) .. letting them in here though

        abline(v=log10(c(min.cell.size,max.cell.size)),lty=2,col=2)
        # look at the number of genes vs. molecule size depenency
        df <- data.frame(molecules=colSums(countMatrix),genes=colSums(countMatrix>0));
        df <- df[df$molecules>=min.cell.size,];
        df <- log10(df);
        df <- df[order(df$molecules,decreasing=F),]
        plot(df,col=adjustcolor(1,alpha=alpha),cex=0.5,ylab='log10[ gene counts]',xlab='log10[ molecule counts]')
        abline(v=log10(c(min.cell.size,max.cell.size)),lty=2,col=2)
        #abline(lm(genes ~ molecules, data=df),col=4)
        require(MASS)
        m <- rlm(genes~molecules,data=df)
        suppressWarnings(pb <- data.frame(predict(m,interval='prediction',level = 1-p.level,type="response")))
        polygon(c(df$molecules,rev(df$molecules)),c(pb$lwr,rev(pb$upr)),col=adjustcolor(2,alpha=0.1),border = NA)
        outliers <- rownames(df)[df$genes > pb$upr | df$genes < pb$lwr];
        points(df[outliers,],col=2,cex=0.6)
        # set of filtered cells to move forward with
        valid.cells <- colSums(countMatrix)>min.cell.size & colSums(countMatrix)<max.cell.size & !(colnames(countMatrix) %in% outliers)
        countMatrix[,valid.cells,drop=F]
}


cd <- t.filter.for.valid.cells(x);
```

Prepare matrix
```{r}
cellPrefix <- '10xPBMC'
cds <- Matrix(cd, sparse=T)
rownames(cds) <- make.unique(rownames(cds))
colnames(cds) <- paste0(cellPrefix, '_', colnames(cds))
```

Perform basic pagoda2 processing
```{r}
# P2 processing
p2 <- Pagoda2$new(cds, n.cores = 20)
p2$adjustVariance(plot=T, gam.k=10)
p2$calculatePcaReduction(nPcs = 100, n.odgenes= 2.e3, maxit= 200)
p2$makeKnnGraph(k = 30, type = 'PCA', center = T, weight.type = 'none', n.cores = 30, distance = 'cosine')
p2$getKnnClusters(method = infomap.community, type = 'PCA', name='infomap')
p2$getKnnClusters(method = multilevel.community, type = 'PCA', name='multilevel')
p2$getKnnClusters(method = walktrap.community, type = 'PCA', name='walktrap')

# Largevis embedding
M <- 30
p2$getEmbedding(type = 'PCA', embeddingType = 'largeVis', M = M, perplexity = 30, gamma = 1 / M, alpha = 1 )

# tSNE embedding
p2$getEmbedding(type = 'PCA', embeddingType = 'tSNE', perplexity = 50)
```

Generate a GO environment for the organism of interest. Functions are available for human, mouse and zebrafish (p2.generate.human.go, p2.generate.mouse.go and p2.generate.dr.go respectively)
```{r}
require(org.Hs.eg.db)
require(GO.db)
# Generate the GO annotation ( in the case for human )
go.env <- p2.generate.human.go(p2)
```

Test path way overdispersion for the GO annotation generated above. This process can take a long time for large datasets
```{r}
p2$testPathwayOverdispersion(
      setenv = go.env,
      verbose = T,
      correlation.distance.threshold = 0.8, # @ 0.9 0.1 correlated get folded together
      recalculate.pca = F,
      min.pathway.size = 50,
      max.pathway.size = 1000)
```

Generate a go environment for the web app
```{r}
geneSets <- p2.generate.human.go.web(colnames(p2$counts),n.cores=4)
```

# Generate the web app

Setup the metadata
```{r}
# Make metadata
additionalMetadata <- list()
additionalMetadata$infomap <- p2.metadata.from.factor(p2$clusters$PCA$infomap, displayname = 'Infomap', s = 0.9)
additionalMetadata$multilevel <- p2.metadata.from.factor(p2$clusters$PCA$multilevel, displayname = 'Multilevel', s = 0.8)
additionalMetadata$walktrap <- p2.metadata.from.factor(p2$clusters$PCA$walktrap, displayname = 'Walktrap', s = 0.5)
```

Generate gene knn graph for searching genes on the web app
```{r}
p2$makeGeneKnnGraph()
```

Generate web app
```{r}
# Generate the web application
# dendrogramCellGroups specifies the clustering that will be
# used to generate the main dendrogram
myPagoda2WebObject <- make.p2.app(p2, dendrogramCellGroups = p2$clusters$PCA$infomap, additionalMetadata = additionalMetadata, geneSets = geneSets);
```

# Optional showing of app
```{r}
show.app(myPagoda2WebObject, name='newPagoda')
```

# Save serialised web object, RDS app and session image
```{r}
myPagoda2WebObject$serialiseToStatic(text.file.directory = './tmp', binary.filename = '10xPBMC.bin')
saveRDS(myPagoda2WebObject, paste0(appName, '.rds'));
save.image.fast('standardAnalysisFinal.RDataF');
```