---
title: "Pagoda2.1 Single-Dataset Workflow"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Pagoda2.1 Single-Dataset Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)

library(pagoda2)
library(ggplot2)
```

## Overview

This vignette shows the standard pagoda2.1 workflow for one single-cell RNA-seq
dataset. The input is a 10x-style count matrix with genes in rows and cells in
columns. The output is a processed `Pagoda2` object with QC metrics, PCA, a
nearest-neighbor graph, UMAP coordinates, Leiden clusters, marker genes, and
publication-ready summary plots.

Pagoda2.1 is designed so the common path is short:

```r
p2 <- Pagoda2$from10x(data_dir)
p2$runQC()
p2$plotQC()
p2$filterCells()
p2$run()
```

The sections below run those steps and explain what each one adds to the object.

## Data

The example uses the first sample from GSE192391, `GSM5746259`. Set
`PAGODA2_GSM5746259_10X` if the triplet files are stored somewhere else.

```{r data-path}
sample_dir <- Sys.getenv("PAGODA2_GSM5746259_10X", unset = "")
if (!dir.exists(sample_dir)) {
  candidates <- c(
    file.path("..", "..", "tests", "data", "GSE192391", "GSM5746259_MGI0369_1_SLAB-145-0"),
    file.path("..", "tests", "data", "GSE192391", "GSM5746259_MGI0369_1_SLAB-145-0"),
    file.path("tests", "data", "GSE192391", "GSM5746259_MGI0369_1_SLAB-145-0")
  )
  sample_dir <- candidates[dir.exists(candidates)][1]
}
stopifnot(dir.exists(sample_dir))
```

Create the pagoda2 object directly from the sample directory:

```{r read}
p2 <- Pagoda2$from10x(sample_dir)

counts <- p2$getRawCounts()
c(cells = nrow(counts), genes = ncol(counts))
```

`Pagoda2$from10x()` reads the count triplet, keeps raw counts in the object, and
sets up the default normalized expression view used by downstream methods.

## Quality Control

The QC step records molecule counts, detected-gene counts, and a default
gene-versus-molecule pass/fail decision in `p2$cellMeta`.

```{r qc}
p2$runQC()
head(p2$getCellMeta(c("n_molecules", "n_genes", "qc_pass"), resolved = TRUE))
```

The QC plot shows the relationship between molecule counts and detected genes.
Cells flagged by the default rule are shown in red.

```{r qc-plot, fig.height=5}
p2$plotQC()
```

Filter the current object to cells that pass QC:

```{r qc-filter}
p2$filterCells()

c(cells = nrow(p2$getRawCounts()), genes = ncol(p2$getRawCounts()))
```

If you skip `runQC()`, `filterCells()` will run QC first. If you skip both and
call `p2$run()`, pagoda2 will calculate QC metrics and warn if cells appear to
need filtering, but it will not remove cells unless you ask explicitly.

## Run The Default Workflow

The default run performs variance modeling, PCA, graph construction, UMAP,
Leiden clustering, and marker detection. Results are stored under canonical
names so later plotting calls can usually omit selectors.

```{r run}
p2$run(plots = "none")
```

After the run, the default grouping is the Leiden clustering:

```{r groups}
p2$getDefaultGrouping()
table(p2$getGrouping())
```

## PCA Elbow Plot

The PCA elbow plot helps show how much of the total PCA-input variance is
captured by each component. The solid curve shows the percent explained by each
component; the dashed curve shows the cumulative amount explained.

```{r pca-elbow, fig.height=4}
p2$plotPCAElbow()
```

## UMAP

`plotEmbedding()` uses the default PCA reduction, default UMAP embedding, and
default grouping unless you specify otherwise.

```{r umap, fig.height=6}
p2$plotEmbedding(
  mark.groups = TRUE,
  size = 0.35,
  alpha = 0.55,
  title = "GSM5746259 Leiden clusters"
)
```

## Marker Genes

The default run stores marker genes for the default grouping. A dot plot gives a
compact summary of marker expression by cluster.

```{r marker-dotplot, fig.width=12, fig.height=7}
p2$plotMarkerDotPlot() +
  labs(title = "Leiden marker genes")
```

The native marker heatmap uses the same marker result and grouping. Here the
top annotation tracks show the QC metrics calculated earlier.

```{r marker-heatmap, fig.width=12, fig.height=8}
p2$plotMarkerHeatmap(
  column.metadata = c("n_molecules", "n_genes"),
  split = TRUE,
  show_heatmap_legend = TRUE
)
```

## What To Do Next

For real analyses, the next step is usually biological annotation. You can store
cluster annotations as another cell metadata column and make it the default
grouping:

```r
p2$annotateClusters(
  from = "leiden",
  to = "cell_type",
  map = c("1" = "T cells", "2" = "monocytes"),
  setDefault = TRUE
)
```

After that, the same plotting and marker methods will use `cell_type` by default
unless you request a different grouping.
