---
title: "Pagoda2.1 Multiome Workflow (RNA + ATAC)"
format:
  html:
    toc: true
    toc-depth: 3
    self-contained: true
  ipynb: default
jupyter: ir44
---

pagoda2.1 represents a multi-modal dataset as a set of **facets** — one per molecular modality measured in the same cells. A facet bundles its modality's raw counts, the normalization *view recipe* suited to that modality, and its reductions; all facets share the same cells. Loading a plain count matrix creates a single **default facet** named `RNA` (so ordinary scRNA-seq needs no facet arguments — see the [single-dataset notebook](pagoda2.1-single-dataset.ipynb)); extra modalities become extra facets, and the analysis verbs take `facet =` (one modality) or `facets =` (integrate across them). The same verbs cover every modality — these are illustrative, not run here:

```r
p2$listFacets()                                  # which modalities are present, e.g. "RNA", "ATAC"
p2$runReduction(facet = "ATAC", method = "lsi")  # reduce one modality (LSI for ATAC)
p2$runGraph(facets = c("RNA", "ATAC"), method = "wnn")  # integrate across modalities
p2$plotEmbedding(gene = "CD3D", facet = "RNA")   # color the embedding by a facet's feature
```

Steps written without `facet =` act on the default `RNA` facet.

This notebook uses a 10x [multiome](https://www.10xgenomics.com/datasets/pbmc-from-a-healthy-donor-granulocytes-removed-through-cell-sorting-3-k-1-standard-2-0-0) sample (PBMC, granulocyte-sorted 3k): each cell is measured by both gene expression (RNA) and chromatin accessibility (ATAC peaks). The two modalities need different treatment — RNA is reduced by PCA, ATAC by LSI — and are then integrated with **weighted nearest neighbors (WNN)**. The pattern is the same as the [CITE-seq notebook](pagoda2.1-citeseq.ipynb); only the per-facet reductions differ. The 10x file is in a local folder named `data`.

## Import: one object, two facets

A multiome 10x HDF5 carries both `Gene Expression` and `Peaks` feature types; pagoda2 reads the per-feature `feature_type` and builds an `RNA` and an `ATAC` facet automatically, each with the right view model. RNA uses `plain` (library-size normalize, then `log1p`). ATAC uses `tfidf` (term-frequency × inverse-document-frequency): accessibility is near-binary and there are tens of thousands of peaks, so TF-IDF down-weights ubiquitous, uninformative peaks and up-weights cell-type-specific ones.

```{r}
p2 <- pagoda2::Pagoda2$from10xH5("data/pbmc_granulocyte_sorted_3k_filtered_feature_bc_matrix.h5", verbose = FALSE)
sapply(p2$listFacets(), function(f) p2$getFacet(f)$modelType)   # facet -> view model
```

`RNA` is the default facet (steps without a `facet=` act on it); `ATAC` adds tens of thousands of peaks over the same cells.

## QC

```{r}
p2$runQC(verbose = FALSE)
```

## Reduce each modality on its own terms

The pipeline verbs are generic — the algorithm is a `method=`. For RNA we run PCA on the overdispersed genes. For ATAC we run **LSI** (latent semantic indexing): SVD of the TF-IDF view, dropping the first component, which tracks sequencing depth rather than biology. LSI is to ATAC what PCA is to RNA. Here we run LSI on the most-accessible peaks for speed; drop the `genes =` subset for a full analysis. Scores are name-keyed: `reductions[["PCA"]]` for RNA, `reductions[["ATAC:LSI"]]` for ATAC.

```{r}
p2$runVariance(verbose = FALSE)
p2$runReduction(nPcs = 30, verbose = FALSE)
top.peaks <- names(sort(Matrix::colSums(p2$getFacet("ATAC")$rawCounts), decreasing = TRUE))[1:20000]
p2$runReduction(facet = "ATAC", method = "lsi", genes = top.peaks, nPcs = 30, verbose = FALSE)
names(p2$reductions)
```

The default `RNA` facet also has a one-call shortcut for its standard sequence (QC → filter → variance → PCA → graph → UMAP → Leiden → markers), which you can run whole or in part:

```r
p2$run()                                  # full standard pipeline on the RNA facet
p2$run(steps = c("variance", "pca"))      # just the RNA reduction
```

We run the per-facet steps explicitly here so RNA and ATAC each get the reduction they need before WNN integration.

## Integrate the modalities with WNN

**WNN** (weighted nearest neighbors; Hao et al., *Cell* 2021) learns, for each cell, how informative each modality's neighborhood is about that cell, turns it into a **per-cell weight per modality**, and fuses the per-modality neighbor graphs accordingly — local weighting, not a single global ratio. `runGraph(method = "wnn")` stores the per-cell weights (`cellMeta$wnn_weight_RNA`, `..._ATAC`), the fused graph (`graphs[["WNN"]]`), and a joint reduction (`reductions[["WNN"]]`); we then cluster on the fused graph with Leiden.

```{r}
p2$runGraph(method = "wnn", facets = c("RNA", "ATAC"), verbose = FALSE)
p2$runClustering(graph = "WNN", name = "wnn_leiden")
```

## What each modality resolves

To see what integration buys, embed each facet's own reduction and the joint WNN reduction, and color all three by the **same** joint clusters. Where a panel mixes cluster colors, that modality alone cannot separate those cells; the WNN panel shows the combined resolution.

```{r}
p2$runEmbedding(reduction = "PCA", name = "umap", verbose = FALSE)        # RNA alone
p2$runEmbedding(reduction = "ATAC:LSI", name = "umap", verbose = FALSE)   # ATAC alone
p2$runEmbedding(reduction = "WNN", name = "umap", verbose = FALSE)        # joint
```

```{r}
#| fig-width: 13
#| fig-height: 4.3
#| fig-dpi: 110
g_rna <- p2$plotEmbedding(reduction = "PCA", embedding = "umap", grouping = "wnn_leiden", mark.groups = TRUE, size = 0.4, alpha = 0.5, title = "RNA only")
g_atac <- p2$plotEmbedding(reduction = "ATAC:LSI", embedding = "umap", grouping = "wnn_leiden", mark.groups = TRUE, size = 0.4, alpha = 0.5, title = "ATAC only")
g_wnn <- p2$plotEmbedding(reduction = "WNN", embedding = "umap", grouping = "wnn_leiden", mark.groups = TRUE, size = 0.4, alpha = 0.5, title = "WNN joint")
cowplot::plot_grid(g_rna, g_atac, g_wnn, nrow = 1)
```

The per-cell ATAC weight maps where chromatin accessibility, rather than RNA, was the more informative modality:

```{r}
#| fig-width: 5
#| fig-height: 4.5
#| fig-dpi: 110
p2$plotEmbedding(reduction = "WNN", embedding = "umap",
                 colors = stats::setNames(p2$cellMeta$wnn_weight_ATAC, rownames(p2$cellMeta)),
                 size = 0.5, alpha = 0.7, title = "per-cell ATAC weight")
```

## Markers

Genes are the interpretable modality, so we call markers on the RNA facet for the joint clusters (ATAC contributes accessibility structure to the integration, but cluster identity is read from gene expression). The dot plot shows the top markers per cluster — color is mean expression, dot size the fraction of cells expressing.

```{r}
p2$runMarkers(grouping = "wnn_leiden", name = "wnn_markers", verbose = FALSE)
```

```{r}
#| fig-width: 14
#| fig-height: 7.5
#| fig-dpi: 130
p2$plotMarkerDotPlot(markers = "wnn_markers", n.genes.per.group = 3, selection = "balanced",
                     order.groups = TRUE, dot.scale = 6) +
  ggplot2::labs(title = "RNA markers by WNN cluster") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Canonical lineage genes on the joint embedding confirm the clusters — `plotEmbedding(gene = ...)` colors by a gene's expression (the default RNA facet here):

```{r}
#| fig-width: 11
#| fig-height: 3.8
#| fig-dpi: 110
genes <- intersect(c("CD3D", "MS4A1", "CD14"), colnames(p2$getRawCounts()))   # T / B / monocyte
plots <- lapply(genes, function(g)
  p2$plotEmbedding(reduction = "WNN", embedding = "umap", gene = g, size = 0.5, alpha = 0.7, title = g))
cowplot::plot_grid(plotlist = plots, nrow = 1)
```

## Next steps

Annotate the WNN clusters from the RNA markers (`p2$annotateClusters(from = "wnn_leiden", ...)`). Reduction-level integration is also available (`p2$runReduction(facets = c("RNA","ATAC"), method = "cca")`), and a disk-backed `backend = "lstar"` facet keeps the large peak matrix out of memory for atlas-scale data. See the [single-dataset](pagoda2.1-single-dataset.ipynb) (RNA) and [CITE-seq](pagoda2.1-citeseq.ipynb) (RNA + protein) notebooks for the other modalities.
