---
title: "Pagoda2.1 CITE-seq Workflow (RNA + protein)"
format:
  html:
    toc: true
    toc-depth: 3
    self-contained: true
  ipynb: default
jupyter: ir44
---

pagoda2.1 can analyze datasets with **more than one molecular modality** measured in the same cells — CITE-seq (gene expression + surface-protein antibody counts), 10x multiome (gene expression + chromatin accessibility), and so on. Each modality is represented as a **facet**.

A facet bundles everything specific to one modality: its raw count matrix, the normalization *view recipe* appropriate for that modality, and the reductions computed from it. All facets share the same cells. When you load an ordinary count matrix, pagoda2 creates a single **default facet** named `RNA` — so a plain scRNA-seq analysis (the [single-dataset notebook](pagoda2.1-single-dataset.ipynb)) never mentions facets at all. Additional modalities become additional facets, and the analysis verbs gain a `facet =` argument (act on one modality) or a `facets =` argument (integrate across them). The shape of the API is the same throughout — these are illustrative, not run here:

```r
p2$listFacets()                                  # which modalities are present, e.g. "RNA", "ADT"
p2$runReduction(facet = "ADT")                   # reduce one modality
p2$runGraph(facets = c("RNA", "ADT"), method = "wnn")   # integrate across modalities
p2$runMarkers(grouping = "leiden", facet = "ADT")       # call markers on one modality
p2$getExpressionBlock(facet = "ADT")             # pull one facet's normalized values
```

Steps written without `facet =` act on the default `RNA` facet, which is why single-modality analysis never needs the argument.

This notebook uses a CITE-seq sample — 10x [PBMC 1k protein v3](https://www.10xgenomics.com/datasets/1-k-pbm-cs-using-3-prime-feature-barcoding-3-standard-3-0-0) — which carries RNA ("Gene Expression") and a 17-antibody surface-protein panel ("Antibody Capture"). We reduce each modality on its own terms and then integrate them with **weighted nearest neighbors (WNN)** into a single joint clustering and embedding. The 10x file is in a local folder named `data`.

## Import: one object, two facets

A multimodal 10x HDF5 records a `feature_type` for every feature; pagoda2 reads it and builds one facet per modality automatically. Each facet gets the normalization view model appropriate for its modality — RNA uses `plain` (library-size normalize, then `log1p`); CITE-seq protein uses `clr`, the centered log-ratio across the antibody panel within each cell, which removes per-cell differences in total antibody capture (a small, compositional panel where raw counts are not comparable across cells).

```{r}
p2 <- pagoda2::Pagoda2$from10xH5("data/pbmc_1k_protein_v3_filtered_feature_bc_matrix.h5", verbose = FALSE)
sapply(p2$listFacets(), function(f) p2$getFacet(f)$modelType)   # facet -> view model
```

Both facets cover the same cells. `RNA` is the default facet, so steps written without a `facet=` argument act on it.

## QC

`runQC()` computes per-cell quality metrics on the RNA facet — molecule and detected-gene counts and a gene-vs-molecule pass/fail flag — and records them in `p2$cellMeta`.

```{r}
p2$runQC(verbose = FALSE)
```

## Reduce each modality on its own terms

The pipeline verbs are generic: the algorithm is chosen by `method=`, so the same `runReduction()` does PCA on RNA and (in the multiome notebook) LSI on ATAC — there is no `runPCA`/`runLeiden`. Here we model gene variance and run PCA on the overdispersed RNA genes, and separately run PCA on the CLR-transformed protein panel. Reduction scores are stored under name-keyed slots — the default facet's is `reductions[["PCA"]]`, the ADT facet's is `reductions[["ADT:PCA"]]`.

```{r}
p2$runVariance(verbose = FALSE)
p2$runReduction(nPcs = 25, verbose = FALSE)
p2$runVariance(facet = "ADT", use.raw.variance = TRUE, verbose = FALSE)
p2$runReduction(facet = "ADT", nPcs = 15, verbose = FALSE)
names(p2$reductions)
```

For the default `RNA` facet the whole standard sequence (QC → filter → variance → PCA → graph → UMAP → Leiden → markers) chains in a single call — `p2$run()` — and you can restrict it to a few steps:

```r
p2$run()                                       # full standard pipeline on the RNA facet
p2$run(steps = c("qc", "variance", "pca"))     # just the reduction steps
```

We process each modality step by step here so the protein panel gets its own variance and reduction settings before the two are integrated.

## Integrate the modalities with WNN

With a reduction per modality, we integrate. **WNN** (weighted nearest neighbors; [Hao et al., *Cell* 2021](https://doi.org/10.1016/j.cell.2021.04.048)) asks, for each cell, how well its neighbors in each modality predict that cell's own profile, and turns the answer into a **per-cell weight for each modality**. A cell whose protein neighbors are more predictive than its RNA neighbors leans on protein, and vice versa — the integration is local, not a single global mixing ratio. WNN then fuses the per-modality neighbor graphs weighted this way into one graph.

`runGraph(method = "wnn")` produces the per-cell weights (`cellMeta$wnn_weight_RNA`, `..._ADT`), the fused graph (`graphs[["WNN"]]`), and a joint reduction for embedding (`reductions[["WNN"]]`). We then cluster on the fused graph with Leiden.

```{r}
p2$runGraph(method = "wnn", facets = c("RNA", "ADT"), verbose = FALSE)
p2$runClustering(graph = "WNN", name = "wnn_leiden")
```

## What each modality resolves

Why integrate at all? Because different populations are best distinguished by different modalities. To see this directly, we embed each facet's own reduction *and* the joint WNN reduction, then color all three UMAPs by the **same** joint clusters. Where a panel mixes cluster colors together, that modality alone cannot separate those cells; where the WNN panel pulls them apart, integration recovered the distinction.

```{r}
p2$runEmbedding(reduction = "PCA", name = "umap", verbose = FALSE)        # RNA alone
p2$runEmbedding(reduction = "ADT:PCA", name = "umap", verbose = FALSE)    # protein alone
p2$runEmbedding(reduction = "WNN", name = "umap", verbose = FALSE)        # joint
```

```{r}
#| fig-width: 13
#| fig-height: 4.3
#| fig-dpi: 110
g_rna <- p2$plotEmbedding(reduction = "PCA", embedding = "umap", grouping = "wnn_leiden", mark.groups = TRUE, size = 0.4, alpha = 0.5, title = "RNA only")
g_adt <- p2$plotEmbedding(reduction = "ADT:PCA", embedding = "umap", grouping = "wnn_leiden", mark.groups = TRUE, size = 0.4, alpha = 0.5, title = "protein only")
g_wnn <- p2$plotEmbedding(reduction = "WNN", embedding = "umap", grouping = "wnn_leiden", mark.groups = TRUE, size = 0.4, alpha = 0.5, title = "WNN joint")
cowplot::plot_grid(g_rna, g_adt, g_wnn, nrow = 1)
```

RNA alone tends to merge the T-cell subsets; the protein panel separates them (by CD4/CD8/CD45RA/RO) but is coarser elsewhere; the WNN joint embedding combines the two. The per-cell ADT weight below maps where protein, rather than RNA, was the more informative modality — typically the protein-defined immune subsets.

```{r}
#| fig-width: 5
#| fig-height: 4.5
#| fig-dpi: 110
p2$plotEmbedding(reduction = "WNN", embedding = "umap",
                 colors = stats::setNames(p2$cellMeta$wnn_weight_ADT, rownames(p2$cellMeta)),
                 size = 0.5, alpha = 0.7, title = "per-cell ADT weight")
```

## Markers

We call marker genes on the RNA facet (the directly interpretable modality) for the joint clusters. The dot plot shows the top markers per cluster — color is mean expression, dot size is the fraction of cells expressing the gene.

```{r}
p2$runMarkers(grouping = "wnn_leiden", name = "wnn_markers", verbose = FALSE)
```

```{r}
#| fig-width: 14
#| fig-height: 7.5
#| fig-dpi: 130
p2$plotMarkerDotPlot(markers = "wnn_markers", n.genes.per.group = 3, selection = "balanced",
                     order.groups = TRUE, dot.scale = 6) +
  ggplot2::labs(title = "RNA markers by WNN cluster") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5))
```

On the protein side every antibody is detected in nearly every cell — that is exactly what CLR normalizes — so a dot plot's "fraction expressing" dimension carries no information. The clearer summary is the mean CLR of each protein across clusters (`getExpressionBlock(facet = "ADT")` returns the CLR matrix). The canonical surface markers light up the joint clusters: CD3/CD4/CD8 on T cells, CD19 on B cells, CD14/CD15 on monocytes, CD16/CD56 on NK.

```{r}
#| fig-width: 8.5
#| fig-height: 5.5
#| fig-dpi: 130
adt <- as.matrix(p2$getExpressionBlock(facet = "ADT"))                                  # cells x proteins, CLR
prot <- t(sapply(split(rownames(adt), p2$cellMeta[rownames(adt), "wnn_leiden"]),        # mean per cluster
                 function(cells) colMeans(adt[cells, , drop = FALSE])))
colnames(prot) <- sub("_TotalSeqB$", "", colnames(prot))
z <- scale(prot)                                                                        # z-score per protein
df <- data.frame(cluster = rownames(z)[row(z)], protein = factor(colnames(z)[col(z)], colnames(z)), z = c(z))
ggplot2::ggplot(df, ggplot2::aes(protein, cluster, fill = z)) +
  ggplot2::geom_tile() +
  ggplot2::scale_fill_gradient2(low = "#2166ac", mid = "grey97", high = "#b2182b", name = "z(CLR)") +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1), panel.grid = ggplot2::element_blank()) +
  ggplot2::labs(title = "Mean surface-protein (CLR) by WNN cluster", x = NULL, y = "cluster")
```

## Next steps

The RNA and protein evidence together name each WNN cluster; record the annotation and make it the default grouping with `p2$annotateClusters(from = "wnn_leiden", to = "cell_type", map = ..., setDefault = TRUE)`, after which every plot uses it. Other integration choices exist at the reduction level (`p2$runReduction(facets = c("RNA","ADT"), method = "cca")`), and any feature from any facet can color the embedding (`p2$plotEmbedding(gene = "CD3_TotalSeqB", facet = "ADT")`). See the [single-dataset](pagoda2.1-single-dataset.ipynb) (RNA) and [ATAC / multiome](pagoda2.1-multiome.ipynb) (RNA + chromatin) notebooks for the other modalities.
