---
title: "Pagoda2.1 Single-Dataset Workflow"
format:
  html:
    toc: true
    toc-depth: 3
    self-contained: true
  ipynb: default
jupyter: ir44
---

This notebook shows the current pagoda2.1 single-dataset workflow on a PBMC sample from GEO: [GSM5746259](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM5746259).

The count data are downloaded from that GEO sample page as a 10x-style triplet: a sparse matrix file, a cell barcode file, and a gene/feature file. Here those three files are in a local folder named `data`.

## Input Data

The folder contains the three files needed for a 10x-style sparse count matrix. The next cell shows the folder name and the files that were downloaded from GEO.

```{r}
data_dir <- "data"

data.frame(
  data_dir = data_dir,
  file = list.files(data_dir, pattern = "\\.(mtx|tsv)(\\.gz)?$")
)
```

Create the pagoda2 object directly from that folder. The reader keeps raw counts in the object and creates a lightweight normalized analysis view for downstream calculations.

```{r}
p2 <- suppressMessages(pagoda2::Pagoda2$from10x(data_dir, verbose = FALSE))
raw_dim <- dim(p2$getRawCounts())
raw_nnz <- Matrix::nnzero(p2$getRawCounts())

data.frame(
  cells = raw_dim[1],
  genes = raw_dim[2],
  nonzero_entries = raw_nnz,
  sparsity = round(1 - raw_nnz / prod(raw_dim), 4),
  integer_like = all(abs(p2$getRawCounts()@x - round(p2$getRawCounts()@x)) < 1e-8)
)
```

## QC

`runQC()` records molecule counts, detected-gene counts, and a gene-versus-molecule QC decision in `p2$cellMeta`. Mitochondrial and ribosomal percentages are included when the gene annotation supports common prefix matching.

```{r}
suppressMessages(invisible(p2$runQC(verbose = FALSE)))
qc <- p2$resolveCellMeta(c("n_molecules", "n_genes", "qc_pass"))

data.frame(
  cells = nrow(qc),
  qc_pass = sum(as.logical(qc$qc_pass), na.rm = TRUE),
  qc_fail = sum(!as.logical(qc$qc_pass), na.rm = TRUE),
  fail_fraction = round(mean(!as.logical(qc$qc_pass), na.rm = TRUE), 4),
  median_molecules = median(qc$n_molecules, na.rm = TRUE),
  median_genes = median(qc$n_genes, na.rm = TRUE)
)
```

The main QC plot shows detected genes versus molecule counts. Cells failing the default QC decision are highlighted in red.

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| fig-dpi: 140
p2$plotQC()
```

Composition QC is annotation-dependent. Here the sample has recognizable mitochondrial and ribosomal gene prefixes, so the violin plot is informative.

```{r}
#| fig-width: 7.5
#| fig-height: 4.5
#| fig-dpi: 140
p2$plotQCViolin()
```

## Run The Default Workflow

The standard run performs filtering, variance modeling, PCA, graph construction, default UMAP embedding, Leiden clustering, and marker detection. Results are stored under canonical names so plotting calls can usually omit selectors.

```{r}
suppressMessages(invisible(p2$run(plots = "none", verbose = FALSE)))

groups <- p2$getGrouping()
summary <- data.frame(
  cells = nrow(p2$getRawCounts()),
  raw_genes = ncol(p2$getRawCounts()),
  analysis_genes = sum(p2$resolveGeneMeta("analysis_pass")$analysis_pass),
  od_genes = length(p2$misc$odgenes),
  default_grouping = p2$getDefaultGrouping(),
  clusters = length(levels(groups))
)
summary
```

```{r}
cluster_sizes <- sort(table(p2$getGrouping()), decreasing = TRUE)
head(data.frame(cluster = names(cluster_sizes), cells = as.integer(cluster_sizes)), 15)
```

## Variance Model QC

`runVariance()` models the mean-variance relationship and selects overdispersed genes for PCA. The diagnostic plot checks the fitted trend and the adjusted variance distribution after normalization. Overdispersed genes are highlighted in red.

```{r}
#| fig-width: 10
#| fig-height: 4.8
#| fig-dpi: 140
p2$plotVarianceQC()
```

## PCA And Default UMAP Embedding

The PCA elbow plot shows percent total variance explained by each PC and cumulatively. Use it to assess whether the default PC range is reasonable before interpreting the graph and embedding.

```{r}
#| fig-width: 7.5
#| fig-height: 4.2
#| fig-dpi: 140
p2$plotPCAElbow()
```

The default UMAP embedding below is colored by the default Leiden clustering. Check for coherent clusters, tiny outlying groups, and possible QC or batch structure before annotating cell types.

```{r}
#| fig-width: 7.4
#| fig-height: 6.2
#| fig-dpi: 140
p2$plotEmbedding(
  mark.groups = TRUE,
  size = 0.35,
  alpha = 0.55,
  title = "GSM5746259 Leiden clusters"
)
```

## Marker Genes

The default run stores marker genes for the default grouping. The dot plot and heatmap use the same marker result and marker-selection logic.

```{r}
marker_name <- "leiden"
marker_name
```

```{r}
#| fig-width: 15
#| fig-height: 10.5
#| fig-dpi: 140
p2$plotMarkerDotPlot(
  markers = marker_name,
  n.genes.per.group = 5,
  selection = "balanced",
  order.groups = TRUE,
  dot.scale = 8.5
) + ggplot2::labs(title = "Leiden marker genes")
```

The native marker heatmap avoids a heavy ComplexHeatmap dependency. The top tracks show QC metrics so marker blocks can be inspected together with per-cell quality summaries.

```{r}
#| fig-width: 13.8
#| fig-height: 9
#| fig-dpi: 140
p2$plotMarkerHeatmap(
  markers = marker_name,
  engine = "native",
  n.genes.per.group = 3,
  selection = "balanced",
  column.metadata = c("n_molecules", "n_genes"),
  row.label.font.size = 9,
  split = TRUE,
  show_heatmap_legend = TRUE
)
```

## Next Steps

For real analyses, the next step is usually biological annotation. Store cluster annotations as another cell metadata column and make it the default grouping when appropriate:

```r
p2$annotateClusters(
  from = "leiden",
  to = "cell_type",
  map = c("1" = "T cells", "2" = "monocytes"),
  setDefault = TRUE
)
```

After that, the same plotting and marker methods will use `cell_type` by default unless a different grouping is requested.
