---
title: "Pagoda2.1 Multiome Workflow (RNA + ATAC)"
format:
  html:
    toc: true
    toc-depth: 3
    self-contained: true
  ipynb: default
jupyter: ir44
---

A 10x [multiome](https://www.10xgenomics.com/datasets/pbmc-from-a-healthy-donor-granulocytes-removed-through-cell-sorting-3-k-1-standard-2-0-0) sample (PBMC, granulocyte-sorted 3k) measures each cell by both gene expression (RNA) and chromatin accessibility (ATAC peaks). Each modality is a **facet**: RNA uses the `plain` view, ATAC uses `tfidf` reduced by **LSI** (TF-IDF → SVD, dropping the depth-correlated first component). The two are integrated with **WNN**, exactly as for CITE-seq — only the per-facet reductions differ. The 10x file is in a local folder named `data`.

## Import

```{r}
p2 <- pagoda2::Pagoda2$from10xH5("data/pbmc_granulocyte_sorted_3k_filtered_feature_bc_matrix.h5", verbose = FALSE)
sapply(p2$listFacets(), function(f) p2$getFacet(f)$modelType)   # facet -> view model
```

## QC

```{r}
p2$runQC(verbose = FALSE)
```

## Per-facet reductions: RNA PCA + ATAC LSI

RNA → PCA on overdispersed genes; ATAC → LSI on its TF-IDF view. For a quick tutorial we run LSI on the most-accessible peaks; drop the `genes =` subset for a full analysis.

```{r}
p2$runVariance(verbose = FALSE)
p2$runReduction(nPcs = 30, verbose = FALSE)
top.peaks <- names(sort(Matrix::colSums(p2$getFacet("ATAC")$rawCounts), decreasing = TRUE))[1:20000]
p2$runReduction(facet = "ATAC", method = "lsi", genes = top.peaks, nPcs = 30, verbose = FALSE)
names(p2$reductions)
```

## WNN integration and clustering

```{r}
p2$runGraph(method = "wnn", facets = c("RNA", "ATAC"), verbose = FALSE)
p2$runClustering(graph = "WNN", name = "wnn_leiden")
```

## What each modality resolves

Embed each facet's reduction and the joint WNN reduction, colored by the **same** joint clusters:

```{r}
p2$runEmbedding(reduction = "PCA", name = "umap", verbose = FALSE)        # RNA alone
p2$runEmbedding(reduction = "ATAC:LSI", name = "umap", verbose = FALSE)   # ATAC alone
p2$runEmbedding(reduction = "WNN", name = "umap", verbose = FALSE)        # joint
```

```{r}
#| fig-width: 13
#| fig-height: 4.3
#| fig-dpi: 110
g_rna <- p2$plotEmbedding(reduction = "PCA", embedding = "umap", grouping = "wnn_leiden", mark.groups = TRUE, size = 0.4, alpha = 0.5, title = "RNA only")
g_atac <- p2$plotEmbedding(reduction = "ATAC:LSI", embedding = "umap", grouping = "wnn_leiden", mark.groups = TRUE, size = 0.4, alpha = 0.5, title = "ATAC only")
g_wnn <- p2$plotEmbedding(reduction = "WNN", embedding = "umap", grouping = "wnn_leiden", mark.groups = TRUE, size = 0.4, alpha = 0.5, title = "WNN joint")
cowplot::plot_grid(g_rna, g_atac, g_wnn, nrow = 1)
```

The per-cell ATAC weight shows where chromatin accessibility drives the joint structure:

```{r}
#| fig-width: 5
#| fig-height: 4.5
#| fig-dpi: 110
p2$plotEmbedding(reduction = "WNN", embedding = "umap",
                 colors = stats::setNames(p2$cellMeta$wnn_weight_ATAC, rownames(p2$cellMeta)),
                 size = 0.5, alpha = 0.7, title = "per-cell ATAC weight")
```

## Markers

Markers are computed on the RNA facet (the interpretable modality); the joint clusters are annotated from these.

```{r}
p2$runMarkers(grouping = "wnn_leiden", name = "wnn_markers", verbose = FALSE)
```

```{r}
#| fig-width: 14
#| fig-height: 7.5
#| fig-dpi: 130
p2$plotMarkerDotPlot(markers = "wnn_markers", n.genes.per.group = 3, selection = "balanced",
                     order.groups = TRUE, dot.scale = 6) +
  ggplot2::labs(title = "RNA markers by WNN cluster") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Canonical lineage genes on the joint embedding confirm the clusters:

```{r}
#| fig-width: 11
#| fig-height: 3.8
#| fig-dpi: 110
genes <- intersect(c("CD3D", "MS4A1", "CD14"), colnames(p2$getRawCounts()))
plots <- lapply(genes, function(g)
  p2$plotEmbedding(reduction = "WNN", embedding = "umap", gene = g, size = 0.5, alpha = 0.7, title = g))
cowplot::plot_grid(plotlist = plots, nrow = 1)
```

## Next steps

Annotate the WNN clusters from the RNA markers. Reduction-level integration is also available (`p2$runReduction(facets = c("RNA","ATAC"), method = "cca")`), and a disk-backed `backend = "lstar"` facet keeps large peak matrices out of memory. See the [single-dataset](pagoda2.1-single-dataset.ipynb) and [CITE-seq](pagoda2.1-citeseq.ipynb) notebooks for the other modalities.
